### 2017-07-06 inference.tex

Fig.~\ref{fig:inference-rules} shows the syntax-directed inference
rules for $\txnimp$ commands inside a transaction $\C{TXN}_i$.
Inference depends on the $\I$-constrained rely relation $\R$, and the
high-level invariant $I$ for the reasons that will become clear
shortly.  Inference rules compute, for each command $c$, a (meta)
function $\F$ that returns a set as an expression in $\SL$, given a
pair of sets ($\stl$ and $\stg$) that describe local and a global
database states, respectively. The expression returned by
$\F(\stl,\stg)$ (abstractly) describes the set of records that get
added to $\stl$ as a result of executing $c$ under $\stl$ and $\stg$.
Thus, $\F$ captures the \emph{effect} part of the state transformer of
$c$, which is the function $\lambda(\stl,\stg).~\stl \cup
\F(\stl,\stg)$. For $\F$ to be useful in RG verification, it needs to
be stable w.r.t the rely relation $\R$. The stability condition on
effects is defined below:
\begin{smathpar}
\begin{array}{lcl}
  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg'.~
  \R(\stl,\stg,\stg') \Rightarrow \F(\stl,\stg) \equiv \F(\stl,\stg')
\end{array}
\end{smathpar}
Intuitively, the stability condition requires an effect to describe
the same set of records before and after the interference. The
equivalence in $\SL$ translates to the equality in first-order logic,
as will be explained shortly. In the inference rules, stability is
enforced constructively by a meta function $\stabilize{\cdot}$, which
accepts an effect and returns a new effect that is guaranteed to be
stable under $\R$. $\stabilize{\cdot}$ achieves the stability
guarantee by abstracting away the bound $\stg$ in an unstable $\F$ to
an existentially bound $\stg'$ as described below:
\begin{smathpar}
\begin{array}{lcll}
  \stabilize{\F} & = & \F & \texttt{if } \stable(\R,\F)\\
  & = & \lambda (\stl,\stg).~\existsl(\stg',I(\stg'),\F(\stl,\stg')) & \texttt{otherwise}\\
\end{array}
\end{smathpar}
Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns an
$\F'$ that simply ignores its $\stg$ argument in favor of a generic
$\stg'$, making $\F'$ trivially stable. The existentially bound
$\stg'$ satisfies $I$ since commands always executed against a valid
database state.

@@

@@

\begin{theorem}
if $c \elabsto \F$ and $\semof{\F(\stl,\stg)} = (\phi^1,g)$ and
$P(\stl,\stg) = true$ and $Q(\stl',\stg) = \exists\stl.~ \phi^1 \wedge
  \forall a. a\in\stl' \Leftrightarrow a\in\stl \vee g(a)$, then
  $\R \vdash \hoare{P}{c}{Q}$
\end{theorem}

>>>>

Fig.~\ref{fig:inference-rules} shows the syntax-directed inference
rules for $\txnimp$ commands inside a transaction $\C{TXN}_i$.
Inference depends on the $\I$-constrained rely relation $\R$, and the
high-level invariant $I$ for the reasons that will become clear
shortly.  Inference rules compute, for each command $c$, a (meta)
function $\F$ that returns a set as an expression in $\SL$, given a
pair of sets ($\stl$ and $\stg$) that describe local and a global
database states, respectively. The expression returned by
$\F(\stl,\stg)$ (abstractly) describes the set of records that get
added to $\stl$ as a result of executing $c$ under $\stl$ and $\stg$.
Thus, $\F$ captures the \emph{effect} part of the state transformer of
$c$, which is the function $\lambda(\stl,\stg).~\stl \cup
\F(\stl,\stg)$. For $\F$ to be useful in RG verification, it needs to
be stable w.r.t the rely relation $\R$. The stability condition on
effects is defined below:
\begin{smathpar}
\begin{array}{lcl}
  \stable(\R,\F) & \Leftrightarrow & \forall \stl,\stg,\stg'.~
  \R(\stl,\stg,\stg') \Rightarrow \F(\stl,\stg) \equiv \F(\stl,\stg')
\end{array}
\end{smathpar}
Intuitively, the stability condition requires an effect to describe
the same set of records before and after the interference. The
equivalence in $\SL$ translates to the equality in first-order logic,
as will be explained shortly. In the inference rules, stability is
enforced constructively by a meta function $\stabilize{\cdot}$, which
accepts an effect and returns a new effect that is guaranteed to be
stable under $\R$. $\stabilize{\cdot}$ achieves the stability
guarantee by abstracting away the bound $\stg$ in an unstable $\F$ to
an existentially bound $\stg'$ as described below:
\begin{smathpar}
\begin{array}{lcll}
  \stabilize{\F} & = & \F & \texttt{if } \stable(\R,\F).\\
  & = & \lambda (\stl,\stg).~\existsl(\stg',I(\stg'),\F(\stl,\stg')) 
      & \texttt{otherwise. }\stg'\texttt{ is a fresh name.}\\
\end{array}
\end{smathpar}
Observe that when $\F$ is not stable, then $\stabilize{\F}$ returns an
$\F'$ that simply ignores its $\stg$ argument in favor of a generic
$\stg'$, making $\F'$ trivially stable. The existentially bound
$\stg'$ satisfies $I$ since transactions are always executed against a
valid database state.

@@

% \begin{figure}[h]
% \begin{smathpar}
% \begin{array}{l}
%   \foreache{xs}{(\lambda y.\lambda x.~ 
%         \updatee{(\lambda z.~ \{\idf=z.\idf;\,\C{name}=z.\C{name};\,
%                                \C{s\_id}=x.\idf\})}
%                 {(\lambda z.~ z.\idf = x.\C{s\_id})};\\
%         \hspace*{1.27in} \inserte{x})}
% \end{array}
% \end{smathpar}
% \caption{A $\txnimp$ program that populates a database
%   of couples.}
% \label{fig:people_code}
% \end{figure}

\begin{example}
Consider a database of couples that tracks the spouse
relationship. Each record has an $\idf$ field, a \C{name} field, and a
spouse id (\C{s\_id}) field. A $\txnimp$ transaction bearing id $i$
that populates the database with a given set ($xs$) of couples will
include the following code snippet:
\begin{smathpar}
\begin{array}{l}
  \foreache{xs}{(\lambda y.\lambda x.~ 
        \updatee{(\lambda z.~ \{\idf=z.\idf;\,\C{name}=z.\C{name};\,
                               \C{s\_id}=x.\idf\})}
                {(\lambda z.~ z.\idf = x.\C{s\_id})};\\
        \hspace*{1.27in} \inserte{x})}
\end{array}
\end{smathpar}
The effect inferred for the program, assuming no interference, is as
following:
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~xs \bind 
    (\lambda x. \stg \bind 
      (\lambda z. \itel{z.\idf = x.\C{s\_id}}
                       {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
                               \C{s\_id}=x.\idf\}\}}
                       {\emptyset}) \\
      \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=false; \}\})
\end{array}
\end{smathpar}
Under the possibility of an interference affecting the stability, the
following stable effect is inferred ($I$ is the database invariant):
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~xs \bind 
    (\lambda x.~\existsl(\stg',I,\stg' \bind 
      (\lambda z. \itel{z.\idf = x.\C{s\_id}}
                       {\{\{\idf=z.\idf;\,\C{name}=z.\C{name};\,
                               \C{s\_id}=x.\idf\}\}}
                       {\emptyset})) \\
      \hspace*{1cm} ~\cup~ \{y \,|\, y = \{x \with \delf=false; \}\})
\end{array}
\end{smathpar}
\end{example}

@@

\subsection{Soundness of Inference}

The correspondence between the inference rules and the RG judgment of
Sec.~\ref{sec:reasoning} is stated thus\footnote{Proof can be found in
the supplementary material}:
\begin{theorem}
\label{thm:inference-sound}
  Forall i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
  then $\R \vdash \hoare{{\sf true}}{c}{\F}$
\end{theorem}
Where, ${\sf true} = \lambda(\stl,\stg).~true$ and $\stable(\R,I)
\Leftrightarrow \forall\stl,\stg,\stg'. I(\stg) \conj
\R(\stl,\stg,\stg') \Rightarrow \I(\stg')$.

\subsection{From $\SL$ to the first-order logic}

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by the
inference judgment. The soundness of a transaction's guarantee can now
be established w.r.t the effect $\F$ of the body. The
\rulelabel{RG-Txn} rule updated to use the inference judgment is shown
below:
\begin{smathpar}
\begin{array}{c}
\RULE
{
  \stable(R,\I)\spc
  \stable(R,I)\\
  \R(\stl,\stg,\stg') \Leftrightarrow \I(\C{true},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  c \elabsto \F\\
  \R_c(\stl,\stg,\stg') \Leftrightarrow \I(\C{false},\stl,\stg,\stg') 
        \wedge R(\stg,\stg') \spc
  \stable(\R_c,\F)\\
  \forall \stl,\stg.~ \F(\stl,\stg) \Rightarrow 
    G(\stg, \stl \gg \stg)\spc
  \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
Since the inference judgment $c\elabsto \F$ is fully automatic,
automating the application of the \rulelabel{RG-Txn} rule for a
transaction now requires automating the multiple implication checks in
the premise. While $R$, $\R$, $\R_c$, $\I$ and $\I$ are formulas in
the first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

The algorithm ($\mssemof{\cdot}$) shown in Fig.~\ref{fig:logic}
encodes an $\SL$ expression in a FOL. Given a set expression $s$ of
$\SL$, $\mssemof{s}$ is a pair $(\phi,f)$, where $\phi$ encodes $s$ in
FOL, and $f$ is a boolean function that names the set defined by $s$.
Naming provides an easy handle to $s$, simplifying $\mssemof{\cdot}$.
Set variables ($x$, $\stl$, $\stg$, $\dots$) are encoded as
unconstrained boolean functions ($g_x$, $g_{\stl}$, $g_{\stg}$,
$\ldots$) that are named after the variable.  Encoding of a set
literal $\{a\,|\,\varphi\}$ introduces a new boolean function $g$ that
is $true$ for only those $a$ that satisfy $\varphi$.  Thus $g$
faithfully encodes the literal. The encoding of the $\existsl{}$
expression first eliminates the second-order quantification over sets.
Existentially bound set variables ($a$) is skolemized, i.e.,
instantiated with a new unbound variable $a'$, before encoding the
bound expression $s$ and the first-order constraint $\phi$. The
algorithm for encoding $\phi$, which is already a first-order formula,
is straightforward, hence not shown. It basically involves replacing
set variables with boolean functions named after the variables. The
first-order encoding for a bind expression describes the semantics of
set monad's $\bind$ in FOL. Let $s_1$ be a set, and let $f$ be a
function that maps each variable in $s_1$ to a new set. Then, $s_2 =
s_1 \bind f$ if and only if for all $y\in s_2$, there exists an $x \in
s_1$ such that $y = f(x)$, and forall $x\in s_1$, $f(x)\in s_2$. The
encoding adds new constraints to this effect. Conditional set
expressions and set union are encoded straightforwardly.

### 2017-07-06 reasoning.tex

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It relies on the
transaction-local judgment to verify the transaction body ($c$).
The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_l$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
\I_e(\stl,\stg,\stg')
\end{array}
\end{smathpar}
Thus, $\R_l$ allows an interference only if it does not violate the
execution-time isolation specification ($\I_e$) of the transaction. If a
certain interference violates the isolation spec, i.e., $R(\stg,\stg')
\Rightarrow \neg(\I_e(\stl,\stg,\stg'))$, then $\R_l(\stl,\stg,\stg')
\Leftrightarrow false$, and any assertion is trivially stable w.r.t
that interference. This is sensible considering such interference is
prevented in the operational semantics.

>>>>

\rulelabel{RG-Txn} is the top-level rule that lets us prove a
transaction preserves the high-level invariant $I$ when executed under
the required isolation as specified by $\I$. It relies on the
transaction-local judgment to verify the transaction body ($c$).
The precondition $P$ of $c$ must follow from the fact that the
transaction-local database ($\stl$) is initially empty, and the global
database satisfies the high-level invariant $I$. The rely relation
($\R_l$) for the local judgment is a ternary relation computed as $R$
modulo $\I_e$:
\begin{smathpar}
\begin{array}{lcl}
%\R_l(\stl,\stg,\stg') & \Leftrightarrow & R(\stg,\stg') \conj
%\I_e(\stl,\stg,\stg')
  \R_l(\stl,\stg,\stg') & \Leftrightarrow & \exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg')
\end{array}
\end{smathpar}
In words, if there exists $\stg_1$ which was the global database state during the last step by a transaction and if the transaction can take its next step at the current global database state $\stg$, then even after the interference, the transaction should still be able to take its next step. If there is no such state $\stg_1$, then a transaction will never be able to take its next step after the interference, and hence such interference can be safely discarded.

%\begin{smathpar}
%\exists \stg_1. \I_e(\stl, \stg_1, \stg) \wedge R(\stg, \stg') \wedge \I_e(\stl, \stg_1, \stg') \Rightarrow \forall \stg_2 
%\end{smathpar}
%Thus, $\R_l$ allows an interference only if it does not violate the
%execution-time isolation specification ($\I_e$) of the transaction. If a
%certain interference violates the isolation spec, i.e., $R(\stg,\stg')
%\Rightarrow \neg(\I_e(\stl,\stg,\stg'))$, then $\R_l(\stl,\stg,\stg')
%\Leftrightarrow false$, and any assertion is trivially stable w.r.t
%that interference. This is sensible considering such interference is
%prevented in the operational semantics.
